---
title: "CompareR: Dataset Comparison with CDISC Validation"
author: "Siddharth Lokineni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CompareR: Dataset Comparison with CDISC Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

CompareR is an R package inspired by SAS PROC COMPARE that provides comprehensive tools for comparing datasets with the added power of CDISC validation. The package is designed for data analysts, scientists, and regulatory professionals who need to ensure data quality and compliance with industry standards.

### Key Features

- **Basic Dataset Comparison**: Compare dimensions, variable names, data types, and values
- **Data Preparation**: Sort, filter, clean, and transform datasets before comparison
- **Missing Value Handling**: Multiple strategies for handling missing data
- **Group-wise Comparison**: Compare datasets by grouping variables
- **CDISC Validation**: Validate SDTM and ADaM datasets against regulatory standards
- **Compliance Reports**: Generate detailed HTML and text reports with compliance metrics
- **Flexible Configuration**: Set tolerance levels and comparison options globally

## Getting Started

### Installation

```{r, eval=FALSE}
# Install from CRAN (when available)
install.packages("CompareR")

# Or install from GitHub
devtools::install_github("username/CompareR")
```

### Loading the Package

```{r, eval=FALSE}
library(CompareR)
```

## Basic Dataset Comparison

### Checking Compatibility

Before performing detailed comparisons, it's important to check if two datasets are compatible.

```{r, eval=FALSE}
# Create sample datasets
df1 <- data.frame(
  PatientID = c(1001, 1002, 1003),
  TRT = c("A", "B", "A"),
  Value = c(100.5, 150.2, 120.8)
)

df2 <- data.frame(
  PatientID = c(1001, 1002, 1003),
  TRT = c("A", "B", "A"),
  Value = c(100.5, 150.2, 120.8)
)

# Check if datasets are compatible
compatibility <- CompareR::check_compatibility(df1, df2)
```

### Comparing Variables

Compare variable names and data types between datasets.

```{r, eval=FALSE}
# Compare variables with matching names
df3 <- data.frame(
  PatientID = c(1001, 1002, 1003),
  TRT = c("A", "B", "A"),
  Score = c(100, 150, 120)  # Different column name and type
)

# Identify discrepancies
var_diffs <- CompareR::compare_variables(df1, df3)
# Output shows: different column names (Value vs Score) and type differences
```

### Comparing Observations

Perform row-by-row comparison to identify specific differences.

```{r, eval=FALSE}
# Create datasets with some differences
df4 <- data.frame(
  PatientID = c(1001, 1002, 1003),
  TRT = c("A", "B", "A"),
  Value = c(100.5, 155.2, 120.8)  # Different value in row 2
)

# Compare observations row-by-row
obs_diffs <- CompareR::compare_observations(df1, df4)
# Output identifies differences in row 2, column Value
```

### High-Level Comparison

Get an overview of all differences between datasets.

```{r, eval=FALSE}
# Comprehensive dataset comparison
overview <- CompareR::compare_datasets(df1, df4)
# Returns: dimension comparison, variable discrepancies, data type differences,
# and summary statistics of differences
```

## Data Preparation

### Sorting and Filtering

Prepare datasets by sorting and applying filter criteria.

```{r, eval=FALSE}
# Create datasets with unsorted data
df_unsorted1 <- data.frame(
  Region = c("West", "East", "North", "South"),
  Sales = c(150, 200, 180, 120)
)

df_unsorted2 <- data.frame(
  Region = c("East", "North", "South", "West"),
  Sales = c(210, 185, 125, 160)
)

# Sort both by Region and filter for sales > 150
prepared <- CompareR::prepare_datasets(
  df_unsorted1,
  df_unsorted2,
  sort_columns = "Region",
  filter_criteria = "Sales > 150"
)
# Returns properly aligned and filtered datasets
```

### Cleaning Data

Remove duplicates and standardize text case.

```{r, eval=FALSE}
# Dataset with duplicates and inconsistent case
dirty_data <- data.frame(
  Name = c("Alice", "alice", "Bob", "Bob", "Charlie"),
  Score = c(100, 100, 85, 85, 92)
)

# Clean by removing duplicates and converting to lowercase
clean_data <- CompareR::clean_dataset(
  dirty_data,
  remove_duplicates = TRUE,
  convert_to_case = "lower"
)
# Returns dataset with unique rows and standardized names
```

### Handling Missing Values

Multiple strategies for managing missing data.

```{r, eval=FALSE}
# Dataset with missing values
data_with_na <- data.frame(
  PatientID = c(1001, 1002, 1003, 1004),
  Age = c(45, NA, 62, 51),
  Weight = c(75.5, 80.2, NA, 85.0)
)

# Method 1: Exclude rows with any missing values
excluded <- CompareR::handle_missing_values(data_with_na, method = "exclude")

# Method 2: Replace with specific value
replaced <- CompareR::handle_missing_values(
  data_with_na,
  method = "replace",
  replace_with = list(Age = 50, Weight = 78)
)

# Method 3: Replace with mean
mean_impute <- CompareR::handle_missing_values(data_with_na, method = "mean")

# Method 4: Replace with median
median_impute <- CompareR::handle_missing_values(data_with_na, method = "median")

# Method 5: Flag rows with missing values
flagged <- CompareR::handle_missing_values(data_with_na, method = "flag")
# Adds a new column indicating which rows had missing values
```

### Converting Data Types

Change data types for specific variables.

```{r, eval=FALSE}
# Dataset with incorrect data types
mixed_types <- data.frame(
  PatientID = c("1001", "1002", "1003"),
  VisitDate = c("2023-01-15", "2023-01-16", "2023-01-17"),
  Value = c("100", "150", "120"),
  Flag = c("Y", "N", "Y")
)

# Convert data types
conversions <- list(
  PatientID = "integer",
  VisitDate = "Date",
  Value = "numeric",
  Flag = "factor"
)

converted <- CompareR::convert_data_types(mixed_types, conversions)
# PatientID is now integer, VisitDate is Date, Value is numeric, Flag is factor
```

### Transforming Variables

Apply transformations to create new variables or modify existing ones.

```{r, eval=FALSE}
# Create transformations
transformations <- list(
  Age_Group = function(df) {
    ifelse(df$Age < 30, "Young",
           ifelse(df$Age < 60, "Middle", "Senior"))
  },
  BMI = function(df) {
    (df$Weight / (df$Height^2)) * 10000  # Convert to standard BMI scale
  }
)

# Apply transformations
transformed <- CompareR::transform_variables(data_with_na, transformations)
# New columns Age_Group and BMI are added
```

## Group-wise Comparison

Compare datasets within specific groups or categories.

```{r, eval=FALSE}
# Create datasets with group information
sales_2023 <- data.frame(
  Quarter = c("Q1", "Q2", "Q1", "Q2"),
  Region = c("North", "North", "South", "South"),
  Revenue = c(100000, 120000, 80000, 95000)
)

sales_2024 <- data.frame(
  Quarter = c("Q1", "Q2", "Q1", "Q2"),
  Region = c("North", "North", "South", "South"),
  Revenue = c(115000, 135000, 90000, 105000)
)

# Compare by Region and Quarter
group_comparison <- CompareR::compare_by_group(
  sales_2023,
  sales_2024,
  group_vars = c("Region", "Quarter")
)
# Returns comparison results for each region-quarter combination
```

## Setting Comparison Options

Configure global comparison settings and tolerance levels.

```{r, eval=FALSE}
# Initialize comparison settings
CompareR::initialize_comparison_settings(
  tolerance = 0.01,
  missing_value_method = "exclude"
)

# Set tolerance level for numeric comparisons
CompareR::set_tolerance(0.001)
# Numeric values are considered equal if difference < 0.001

# Reset to default settings
CompareR::reset_comparison_settings()
```

## CDISC Validation (Main Feature)

### What is CDISC?

CDISC (Clinical Data Interchange Standards Consortium) provides standardized formats for regulatory submissions:

- **SDTM** (Study Data Tabulation Model): Raw data from clinical trials
- **ADaM** (Analysis Data Model): Derived datasets used for statistical analysis
- **Compliance**: Required for FDA submissions and regulatory approval

CDISC validation ensures that datasets meet industry standards and regulatory requirements.

### Auto-Detecting CDISC Domain

Automatically identify the domain of a dataset.

```{r, eval=FALSE}
# Create sample SDTM DM dataset (Demographics)
dm_data <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003"),
  SDTM_DOMAIN = "DM",
  AGE = c(45, 62, 51),
  SEX = c("M", "F", "M"),
  RACE = c("WHITE", "BLACK", "ASIAN")
)

# Create sample ADaM ADSL dataset (Subject-Level Analysis)
adsl_data <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003"),
  SAFFL = c("Y", "Y", "Y"),
  ITTFL = c("Y", "Y", "Y"),
  AGE = c(45, 62, 51),
  TRT01P = c("Drug A", "Placebo", "Drug A")
)

# Create non-CDISC dataset
non_cdisc <- data.frame(
  PatientID = c(1, 2, 3),
  Age = c(45, 62, 51),
  Treatment = c("A", "B", "A")
)

# Detect CDISC domains
domain1 <- CompareR::detect_cdisc_domain(dm_data)
# Output: "SDTM: DM (Demographics)"

domain2 <- CompareR::detect_cdisc_domain(adsl_data)
# Output: "ADaM: ADSL (Subject-Level Analysis)"

domain3 <- CompareR::detect_cdisc_domain(non_cdisc)
# Output: "Unknown (Not a standard CDISC dataset)"
```

### Validating SDTM Datasets

Validate SDTM datasets for required variables and compliance.

```{r, eval=FALSE}
# Create SDTM DM dataset with missing required variables
dm_incomplete <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003"),
  # Missing DMDY (day of visit)
  # Missing VISITNUM (visit number)
  AGE = c(45, 62, 51),
  SEX = c("M", "F", "M")
)

# Validate SDTM DM dataset
dm_validation <- CompareR::validate_cdisc(dm_incomplete, domain = "DM", standard = "SDTM")

# Display validation results
CompareR::print_cdisc_validation(dm_validation)
# Output shows:
# ERROR: Missing required variable DMDY
# ERROR: Missing required variable VISITNUM
# WARNING: Missing expected variable RACE
```

### Validating ADaM Datasets

Validate ADaM analysis datasets.

```{r, eval=FALSE}
# Create SDTM AE (Adverse Events) dataset
ae_data <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003"),
  AESTDY = c(1, 5, 3),
  AEENDDY = c(5, 10, 8),
  AEDECOD = c("Headache", "Nausea", "Fever"),
  AESEV = c("MILD", "MODERATE", "MILD"),
  AEACN = c("NONE", "DOSE REDUCED", "NONE")
)

# Validate SDTM AE domain directly
ae_validation <- CompareR::validate_sdtm(ae_data, domain = "AE")

# Create ADaM ADSL (Subject-Level Analysis) dataset
adsl_analysis <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003"),
  SAFFL = c("Y", "Y", "Y"),
  ITTFL = c("Y", "Y", "Y"),
  TRTSDT = c("2023-01-01", "2023-01-05", "2023-01-03"),
  TRT01P = c("Drug A", "Placebo", "Drug A"),
  AGE = c(45, 62, 51)
)

# Validate ADaM ADSL dataset
adsl_validation <- CompareR::validate_adam(adsl_analysis, dataset = "ADSL")
# Checks for required analysis flag variables and treatment assignments
```

### Comparing Datasets with CDISC Compliance

The flagship function that combines comparison with CDISC validation.

```{r, eval=FALSE}
# Create two versions of DM dataset
dm_version1 <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003"),
  SDTM_VERSION = "1.0",
  AGE = c(45, 62, 51),
  SEX = c("M", "F", "M"),
  RACE = c("WHITE", "BLACK", "ASIAN"),
  DMDY = c(1, 1, 1)
)

dm_version2 <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003"),
  SDTM_VERSION = "2.0",
  AGE = c(45, 62, 52),  # Different age for subject 003
  SEX = c("M", "F", "M"),
  RACE = c("WHITE", "BLACK", "ASIAN"),
  DMDY = c(1, 1, 1),
  RACECD = "C" # Additional standardized code
)

# Compare with CDISC compliance checks
cdisc_result <- CompareR::cdisc_compare(
  dm_version1,
  dm_version2,
  domain = "DM",
  standard = "SDTM"
)

# Result structure includes:
# - comparison_results: Row-by-row differences
# - validation_v1: Compliance issues in version 1
# - validation_v2: Compliance issues in version 2
# - differences_summary: Number of differing values by variable
# - compliance_status: Overall pass/fail status
```

### Generating CDISC Reports

Create detailed compliance reports in multiple formats.

```{r, eval=FALSE}
# Generate text report
CompareR::generate_cdisc_report(
  cdisc_result,
  output_format = "text",
  console_output = TRUE
)
# Displays color-coded severity levels:
# [ERROR] - Critical compliance issues
# [WARNING] - Non-critical issues
# [INFO] - Informational messages

# Generate HTML report file
CompareR::generate_cdisc_report(
  cdisc_result,
  output_format = "html",
  file_name = "dm_compliance_report.html"
)
# Creates detailed HTML report with formatted tables and color coding
# Suitable for regulatory submissions and audit trails

# Generate detailed comparison with visual elements
CompareR::generate_cdisc_report(
  cdisc_result,
  output_format = "html",
  file_name = "detailed_comparison.html",
  include_visualizations = TRUE
)
# Includes charts showing variable differences and compliance metrics
```

## Supported CDISC Standards

### SDTM Domains (11 supported)

1. **DM** - Demographics: Subject demographic information
2. **AE** - Adverse Events: Safety data from clinical trials
3. **CM** - Concomitant Medications: Concurrent treatment information
4. **VS** - Vital Signs: Blood pressure, heart rate, temperature
5. **LB** - Laboratory Data: Lab test results and measurements
6. **EX** - Exposure: Drug dosing and administration
7. **MH** - Medical History: Past medical conditions
8. **DS** - Disposition: Subject trial status and reasons
9. **CO** - Comments: Additional narrative information
10. **PC** - Pharmacokinetics: Drug concentration measurements
11. **QS** - Questionnaires: Patient-reported outcomes

### ADaM Datasets (5 supported)

1. **ADSL** - Subject-Level Analysis: Demographics and treatment info
2. **ADAE** - Adverse Events Analysis: Analyzed safety data
3. **ADCM** - Concomitant Medications Analysis: Medication analysis
4. **ADVS** - Vital Signs Analysis: Analyzed vital measurements
5. **ADLB** - Laboratory Analysis: Analyzed lab results

## Reporting

### Summary Reports

Generate high-level summaries of comparisons.

```{r, eval=FALSE}
# Create summary report
summary_report <- CompareR::generate_summary_report(
  df1,
  df2,
  include_missing_values = TRUE,
  include_distributions = FALSE
)
# Output: Dimensions, variable count, data type consistency,
# missing value analysis
```

### Detailed Reports

Create comprehensive reports with all comparison details.

```{r, eval=FALSE}
# Perform comparison
comparison <- CompareR::compare_datasets(df1, df2)

# Generate detailed report
detailed_report <- CompareR::generate_detailed_report(
  comparison,
  output_format = "text",
  include_value_diffs = TRUE,
  include_statistics = TRUE
)
# Shows each difference with context, row numbers, and impact metrics
```

### Visualizations

Create visual representations of comparison results.

```{r, eval=FALSE}
# Generate comparison visualization
visualization <- CompareR::generate_comparison_visualization(
  comparison,
  plot_type = "differences",
  show_missing = TRUE
)

# Save the plot
ggplot2::ggsave("comparison_plot.png", visualization, width = 12, height = 8)
```

## Comparison with SAS PROC COMPARE

| Feature | SAS PROC COMPARE | CompareR |
|---------|-----------------|----------|
| Basic comparison | YES | YES |
| Dimension check | YES | YES |
| Variable comparison | YES | YES |
| Row-wise comparison | YES | YES |
| Data type checking | YES | YES |
| Missing value handling | Limited | Extensive |
| Group-wise comparison | YES | YES |
| Tolerance setting | YES | YES |
| CDISC validation | NO | YES |
| ADaM validation | NO | YES |
| HTML reports | Limited | YES |
| Regulatory compliance | Limited | YES |

## Complete Workflow Example

A comprehensive end-to-end example demonstrating the full power of CompareR with CDISC validation.

```{r, eval=FALSE}
# 1. Load libraries
library(CompareR)
library(dplyr)

# 2. Create two versions of clinical trial demographics
# Baseline version
dm_baseline <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003", "001-004"),
  SDTM_DOMAIN = "DM",
  AGE = c(45, 62, 51, 58),
  SEX = c("M", "F", "M", "F"),
  RACE = c("WHITE", "BLACK", "ASIAN", "WHITE"),
  DMDY = c(1, 1, 1, 1),
  BRTHDTC = c("1978-06-15", "1961-03-22", "1972-09-10", "1965-11-05"),
  ARMCD = c("TRT", "PBO", "TRT", "PBO"),
  ARM = c("Treatment", "Placebo", "Treatment", "Placebo")
)

# Updated version with corrections and additional data
dm_updated <- data.frame(
  USUBJID = c("001-001", "001-002", "001-003", "001-004"),
  SDTM_DOMAIN = "DM",
  SDTM_VERSION = "2.0",
  AGE = c(45, 62, 51, 59),  # Correction: 001-004 age was 58, now 59
  SEX = c("M", "F", "M", "F"),
  RACE = c("WHITE", "BLACK", "ASIAN", "WHITE"),
  DMDY = c(1, 1, 1, 1),
  BRTHDTC = c("1978-06-15", "1961-03-22", "1972-09-10", "1965-11-05"),
  ARMCD = c("TRT", "PBO", "TRT", "PBO"),
  ARM = c("Treatment", "Placebo", "Treatment", "Placebo"),
  RACECD = c("WH", "BL", "AS", "WH")  # Added standardized race codes
)

# 3. Initialize comparison settings
CompareR::initialize_comparison_settings(
  tolerance = 0.001,
  missing_value_method = "exclude"
)

# 4. Check compatibility
cat("Step 1: Checking dataset compatibility...\n")
compat <- CompareR::check_compatibility(dm_baseline, dm_updated)
cat("Compatibility check complete.\n\n")

# 5. Prepare datasets (sort and filter if needed)
cat("Step 2: Preparing datasets...\n")
prepared <- CompareR::prepare_datasets(
  dm_baseline,
  dm_updated,
  sort_columns = "USUBJID"
)
cat("Datasets prepared and sorted.\n\n")

# 6. Perform CDISC-aware comparison
cat("Step 3: Performing CDISC comparison...\n")
cdisc_comparison <- CompareR::cdisc_compare(
  prepared$df1,
  prepared$df2,
  domain = "DM",
  standard = "SDTM"
)
cat("CDISC comparison complete.\n\n")

# 7. Generate detailed comparison results
cat("Step 4: Generating comparison results...\n")
overview <- CompareR::compare_datasets(prepared$df1, prepared$df2)
cat("Comparison overview created.\n\n")

# 8. Generate compliance report
cat("Step 5: Generating compliance report...\n")
CompareR::generate_cdisc_report(
  cdisc_comparison,
  output_format = "html",
  file_name = "dm_compliance_report.html"
)
cat("HTML compliance report generated: dm_compliance_report.html\n\n")

# 9. Generate detailed text report
cat("Step 6: Generating detailed text report...\n")
detailed_report <- CompareR::generate_detailed_report(
  overview,
  output_format = "text"
)
cat(detailed_report)
cat("\n\nDetailed comparison report displayed.\n\n")

# 10. Create visualizations
cat("Step 7: Creating comparison visualizations...\n")
plot <- CompareR::generate_comparison_visualization(
  overview,
  plot_type = "differences"
)
ggplot2::ggsave("dm_comparison_plot.png", plot, width = 12, height = 8)
cat("Visualization saved: dm_comparison_plot.png\n\n")

# 11. Summary
cat("========================================\n")
cat("WORKFLOW COMPLETE\n")
cat("========================================\n")
cat("Generated outputs:\n")
cat("  - dm_compliance_report.html (regulatory compliance)\n")
cat("  - dm_comparison_plot.png (visual analysis)\n")
cat("  - Text report displayed above\n")
cat("\nKey findings:\n")
cat("  - Data quality assessment: COMPLETE\n")
cat("  - CDISC compliance: ",
    if(cdisc_comparison$compliance_status == "PASS") "PASS" else "FAIL", "\n")
cat("  - Differences found in: ",
    length(cdisc_comparison$differences_summary), " variables\n", sep="")
```

## Summary

CompareR combines the power of SAS PROC COMPARE with modern R capabilities and industry-standard CDISC validation. Whether you're performing basic dataset comparisons or ensuring regulatory compliance for clinical trial data, CompareR provides the tools needed for thorough data validation and reporting.

For more information and additional examples, visit the package documentation or GitHub repository.
